Exact Phylodynamics via Structured Markov Genealogy Processes

### 2023-07-19 ms.Rnw
thm 3, appendix

Denote the underlying probability space by $(\Omega,\Borel,\mathrm{Prob})$.
We will assume that our population process is a time-inhomogeneous Markov jump process, $\X_t$, parameterized by time $t\in\Rp\colonequals\{t\in\mathbb{R}\;\vert\;t\ge{0}\}$ and taking values in some space $\Xspace$.
In earlier work \citep{King2022}, we limited ourselves to the case $\Xspace=\Z^d$, but here we assume only that $\Xspace$ is a complete metric space with a countable dense subset, \ie a Polish space.
The population process is completely specified by its initial-state distribution, $p_0$, and its transition rates $\alpha$.
In particular, we suppose that
\begin{equation}\label{eq:ic}
  \Prob{\X_0\in\mathcal{E}}=\int_{\mathcal{E}}{p_0(x)\,\dd{x}}
\end{equation}
for all measurable sets $\mathcal{E}\subseteq\Xspace$.
For any $t\in\Rp$, $x,x'\in\Xspace$, we think of the quantity $\alpha(t,x,x')$ as the instantaneous hazard of a jump from $x$ to $x'$.
More precisely, the transition rates have the following properties:
\begin{equation*}
  \begin{gathered}
    \alpha(t,x,x')\ge{0}, \qquad \int_{\Xspace}{\alpha(t,x,x')\,\dd{x'}}<\infty,\\
  \end{gathered}
\end{equation*}
for all $t\in\Rp$ and $x,x'\in\Xspace$.
Henceforth, we understand that integrals are taken over all of $\Xspace$ unless otherwise specified.
Let $\N_t$ be the number of jumps that $\X$ has taken by time $t$.
We assume that $\N_t$ is a simple counting process so that
\begin{equation*}
  \begin{gathered}
    \Prob{\N_{t+\Delta}=n+1\;\vert\;\N_{t}=n}=\Delta\,\int{\alpha(t,x,x')\,\dd{x'}}+o(\Delta),
    \qquad
    \Prob{\N_{t+\Delta}>n+1\;\vert\;\N_{t}=n}=o(\Delta),\\
    \Prob{\X_{t+\Delta}\in\mathcal{E}\;\vert\;\X_{t}=x, \N_{t+\Delta}-\N_{t}=1}=\frac{\int_{\mathcal{E}}{\alpha(t,x,x')\,\dd{x'}}}{\int{\alpha(t,x,x')\,\dd{x'}}}+\frac{o(\Delta)}{\Delta}.
  \end{gathered}
\end{equation*}
\AAK{[Do we need the last term?]}
We will further assume that $\X_t$ is non-explosive, \ie, that $\Prob{\N_t<\infty}=1$ for all $t$.
\AAK{[Is this equivalent to non-explosivity? Or merely an implication?]}

The above may be compactly summarized by stating that if $w(t,x)$ satisfies the Kolmogorov forward equation (KFE),
%\begin{equation}\label[pluralequation]{eq:kfe}
\begin{equation}\label{eq:kfe}
  \frac{\partial{w}}{\partial{t}}(t,x)=\int\!w(t,x')\,\alpha(t,x',x)\,\dd{x'}-\int\!w(t,x)\,\alpha(t,x,x')\,\dd{x'},
\end{equation}
and if, moreover, $w(0,x) = p_0(x)$,
then $\int_{\mathcal{E}}{w(t,x)\,\dd{x}}=\Prob{\X_t\in\mathcal{E}}$ for every measurable $\mathcal{E}\subseteq{\Xspace}$.
\Cref{eq:kfe} is sometimes called the \emph{master equation} for $\X_t$.

@@

Suppose we have a pruned genealogy $\Prune^*$, defined on the time-interval $[0,T]$, with event times $0=t_0<t_1<\cdots<t_n=T$.
From the theorem, we have, for $t\notin\event(\Prune^*)$,
\begin{equation}
  \frac{\partial{w}}{\partial{t}}=\sum_{u}\int\!w(t,x')\,\alpha_u(t,x',x)\,\phi_u\big(x,y(t),y(t)\big)\,\dd{x'}
  -\sum_{u}\int\!w(t,x)\,\alpha_u(t,x,x')\,\dd{x'}.
\end{equation}
At event times, $t\in\event(\Prune^*)$, one has
\begin{equation}\label{eq:prune_filt_disc}
  w(t,x)=\sum_{u}\int\!\wt(t,x')\,\frac{\alpha_u(t_k,x',x)}{\mu}\,\phi_u\big(x,\yt(t),y(t)\big)\,\dd{x'}\\,
\end{equation}
In addition, there are the initial and boundary conditions
\begin{equation}
  \begin{gathered}
    w(0,x)=p_0(x), \qquad w(t,x)=0\quad\text{whenever}\quad n(x)<\ell(y(t)).
  \end{gathered}
\end{equation}

@@

\appendix

@@

Explicit expressions for the probability densities that will arise in the following are not always available.
Hence, we will develop some technology for manipulating them that avoids the need for explicit expressions.

Suppose $\X_t$ is a continuous-time Markov process with Kolmogorov forward equation (KFE)
\begin{equation}\label{eq:kfe2}
  \frac{\partial{w}}{\partial{t}}
  =\int{w(t,x')\,\beta(t,x',x)\,\dd{x'}}
  -\int{w(t,x)\,\beta(t,x,x')\,\dd{x'}},
\end{equation}
Suppose that $B(t,x,x')>0$ and $\lambda(t,x)\in\R$ are given functions.
We say that the equation
\begin{equation}\label{eq:filtereq}
  \frac{\partial{w}}{\partial{t}}
  =\int{w(t,x')\,\beta(t,x',x)\,B(t,x',x)\,\dd{x'}}
  -\int{w(t,x)\,\beta(t,x,x')\,\dd{x'}}
  -\lambda(t,x)\,w(t,x)
\end{equation}
is the \emph{filter equation} with \emph{driver} $X_t$, \emph{boost} $B$, and \emph{decay} $\lambda$.

Filter equations afford a convenient means of computing the likelihood of a given sequence of events.
This is facilitated by the following fact.

\begin{lemma}\label{lemma:filtereq}
  \Cref{eq:filtereq} is satisfied by $w(t,x)=\int_0^{\infty}{v\,u(t,x,v)\,\dd{v}}$, where $u$ satisfies the KFE
  \begin{equation}\label{eq:filterlemma}
    \begin{aligned}
      \frac{\partial{u}}{\partial{t}}=&\int{u(t,x',v')\,\beta(t,x',x)\,\delta(v,B(t,x',x)\,v')\,\dd{x'}\,\dd{v'}}\\
      &\qquad-\int{u(t,x,v)\,\beta(t,x,x')\,\delta(v',B(t,x,x')\,v)\,\dd{x'}\,\dd{v'}}
      +\tfrac{\partial}{\partial{v}}\left[\lambda(t,x)\,v\,u(t,x,v)\right].
    \end{aligned}
  \end{equation}
  Here, $\delta(v,v')$ is the familiar Dirac $\delta$.
\end{lemma}
\begin{proof}
  \begin{equation*}
    \begin{aligned}
      \frac{\partial{w}}{\partial{t}}
      =&\int{v\,\frac{\partial{u}}{\partial{t}}(t,x,v)\,\dd{v}}\\
      =&\int{v\,u(t,x',v')\,\beta(t,x',x)\,\delta(v,B(t,x',x)v')\,\dd{v}\,\dd{x'}\,\dd{v'}}\\
      &\qquad-\int{v\,u(t,x,v)\,\beta(t,x,x')\,\delta(v',B(t,x,x')v)\,\dd{v}\,\dd{x'}\,\dd{v'}}\\
      &\qquad+\int{v\,\tfrac{\partial}{\partial{v}}\left[\lambda(t,x)\,v\,u(t,x,v)\right]\,\dd{v}}.\\
    \end{aligned}
  \end{equation*}
  Evaluating the first integral with respect to $v$, the second with respect to $v'$, and the third by parts, we obtain
  \begin{equation*}
    \begin{aligned}
      \frac{\partial{w}}{\partial{t}}
      =&\int{v'\,u(t,x',v')\,\beta(t,x',x)\,B(t,x',x)\,\dd{v'}\,\dd{x'}}
      -\int{v\,u(t,x,v)\,\beta(t,x,x')\,\dd{v}\,\dd{x'}}\\
      &\qquad-\lambda(t,x)\,\int{v\,u(t,x,v)\,\dd{v}},
    \end{aligned}
  \end{equation*}
  which is simplified to obtain \cref{eq:filtereq}.
\end{proof}

\Cref{eq:filterlemma} is recognizable as the KFE of a certain process $(X_t,V_t)$.
In particular, $X_t$ is the driver with KFE \cref{eq:kfe2}.
The $V_t$ process has jumps wherever $X_t$ does, such that when $X$ jumps at time $t$ from $x$ to $x'$, $V$ jumps by the multiplicative factor $B(t,x,x')>0$.
Between jumps, $V_t$ decays deterministically and exponentially at rate $\lambda(t,x)$.
If we view $V_t$ as a weight, then \cref{lemma:filtereq} tells us how the $V_t$-weighted average of $X_t$ evolves in time:
this average is simply $\int{w(t,x)\,\dd{x}}$.

This motivates the following result, which shows how filter equations allow one to integrate over random histories.

\begin{prop}\label{prop:zeroboost}
  Suppose $\X_t$ is a continuous-time Markov process with state space $\Xspace$ and KFE as in \cref{eq:kfe2}.
  %% Suppose that $B(t,x,x')>0$, $Q(t,x,x')\in\{0,1\}$, and $\lambda(t,x)\in\R$ are given measurable functions.
  %% Consider the equation
  %% \begin{equation*}
  %%   \frac{\partial{w}}{\partial{t}}
  %%   =\int{w(t,x')\,\beta(t,x',x)\,B(t,x',x)\,Q(t,x',x)\,\dd{x'}}
  %%   -\int{w(t,x)\,\beta(t,x,x')\,\dd{x'}}
  %%   -\lambda(t,x)\,w(t,x)
  %% \end{equation*}
  Let $\Hist_t$ be its history process.
  Suppose $\rdm{R}_t$ is a real-valued random process such that
  \begin{equation*}
    \Expect{\rdm{R}_t\;\vert\;\Hist_t}=\prod_{e\in\event(\Hist_t)}{Q(e,\leftlim{X}_e,X_e)\,B(e,\leftlim{X}_e,X_e)},
  \end{equation*}
  for some given measurable functions $B>{0}$, $Q\in\{0,1\}$.
  Suppose $w$ satisfies the filter equation
  \begin{equation*}
    \begin{aligned}
      \frac{\partial{w}}{\partial{t}}(t,x)
      =&\int{w(t,x')\,\beta(t,x',x)\,Q(t,x',x)\,B(t,x',x)\,\dd{x'}}
      -\int{w(t,x)\,\beta(t,x,x')\,Q(t,x,x')\,\dd{x'}}\\
      &\qquad-\int{w(t,x)\,\beta(t,x,x')\,\Big(1-Q(t,x,x')\Big)\,\dd{x'}}.
    \end{aligned}
  \end{equation*}
  Then $\Expect{\rdm{R}_t}=\int{w(t,x)\dd{x}}$.
\end{prop}
\begin{proof}
  Apply \cref{lemma:filtereq} with the driver generated by the rate functions $\beta(t,x,x')\,Q(t,x,x')$.
\end{proof}

An important special case is that of a deterministic driving process.
The following result is established by routine calculation.

\begin{prop}
  Suppose $X:[0,T]\to\Xspace$ is a deterministic, piecewise constant, \cadlag\ function and let $M=\event(X)$.
  Then the KFE for $X_t$ is \cref{eq:kfe2} with $\beta(t,x,x')=\sum_{e\in{M}}{\delta(t,e)\delta(x',X_e)}$.
  With this driver, the filter equation (\cref{eq:filtereq}) becomes
  \begin{equation*}
    \begin{gathered}
      \frac{\partial{w}}{\partial{t}}=
      -\lambda(t,x)\,w(t,x)\quad\text{for}\quad{t\in[0,T]\setminus{M}},\\
      w(t,x)=\delta(x,X_t)\,\int{\wt(t,x')\,B(t,x',X_t)\,\dd{x'}}\quad\text{for}\quad{t\in{M}}.
    \end{gathered}
  \end{equation*}
\end{prop}

The results so far allow us to compute expectations over random jumps and jump times.
We will have occasion to compute marginal expectations in the situation where some jump times are known.
The following result handles this case.

\begin{prop}
  Suppose $\X_t$ is a continuous-time Markov process with state space $\Xspace$ and KFE
  \begin{equation*}
    \frac{\partial{w}}{\partial{t}}(t,x)=\int\!w(t,x')\,\alpha(t,x',x)\,\dd{x'}-\int\!w(t,x)\,\alpha(t,x,x')\,\dd{x'}.
  \end{equation*}
  Let $\Hist_t$ be its history process.
  Suppose that $0<t_1<\dots<t_N\le{t}$ are fixed times and set $M=\{t_1,\dots,t_N\}$.
  Let $B>{0}$ and $Q\in\{0,1\}$ be given measurable functions.
  Let $\rdm{R}$ be a non-negative real-valued random variable such that
  \begin{equation*}
    \Expect{\rdm{R}\;\vert\;\Hist_t}=\prod_{e\in{\event(\Hist_t)}}{Q(e,\leftlim{X}_e,X_e)\,B(e,\leftlim{X}_e,X_e)}.
  \end{equation*}
  Let $\pi(t,x,x')$ be a given probability kernel and $\Y_t$ be the Markov process generated by \cref{eq:kfe2} where $\beta(t,x,x')=\left[\alpha(t,x,x')+\sum_{e\in{M}}{\delta(t,e)\,\pi(t,x,x')}\right]\,Q(t,x,x')$.
  Suppose $w$ satisfies the filter equation with driver $\Y_t$, boost
  \begin{equation*}
    C(t,x,x')=
    \begin{cases}
      B(t,x,x'), &t\notin{M},\\
      B(t,x,x')\,\displaystyle\frac{\alpha(t,x,x')}{\pi(t,x,x')}, &t\in{M},
    \end{cases}
  \end{equation*}
  and decay $\lambda=\int{\alpha(t,x,x')\,\left(1-Q(t,x,x')\right)\,\dd{x'}}$.
  Then
  \begin{equation*}
    \Expect{\rdm{R}\;\big\vert\;M\subseteq{\event(\Hist_t)}}=\int{w(t,x)\,\dd{x}}.
  \end{equation*}
\end{prop}
\begin{proof}
  Apply \cref{lemma:filtereq}.
\end{proof}

\paragraph{More systematically}

Suppose $\X_t$ is a non-explosive pure jump process with KFE
\begin{equation*}
  \begin{gathered}
    \frac{\partial{w}}{\partial{t}}(t,x)=\int\!w(t,x')\,\alpha(t,x',x)\,\dd{x'}-\int\!w(t,x)\,\alpha(t,x,x')\,\dd{x'},\\
    w(0,x)=p_0(x).
  \end{gathered}
\end{equation*}
We can always take a version of $\X_t$ with \cadlag\ sample paths.

Let $\rdm{\hat{X}}_k$ be the embedded chain of $\X_t$ and let $\rdm{T}^X_k$ be the point process of its jump times.
It is elementary that
\begin{equation*}
  \Prob{\rdm{\hat{X}}_k=\hat{X}_k\;\Big\vert\;\rdm{\hat{X}}_{k-1}=\hat{X}_{k-1},\rdm{T}^X_k=T_k}=\frac{\alpha(T_k,\hat{X}_{k-1},\hat{X}_k)}{\int{\alpha(T_k,\hat{X}_{k-1},x')\dd{x'}}}.
\end{equation*}
Moreover, if one has the $(\rdm{\hat{X}}_k,\rdm{T}^X_k)$ chain, one can reconstruct the $\X_t$ process.
Conditional on $\rdm{\hat{X}}$, the jump-time chain $\rdm{T}^X_k$ is a Poisson point process with intensity $\int{\alpha(t,\X_t,x')\dd{x'}}$.

We can rewrite the KFE as
\begin{equation}\label{eq:poissdriver}
  \begin{aligned}
    \frac{\partial{w}}{\partial{t}}(t,n,x)=&\int\!w(t,n-1,x')\,\nu\,\pi(t,x',x)\,B(t,x,x')\,\dd{x'}-\int\!w(t,n,x)\,\alpha(t,x,x')\,\dd{x'}\\
    =&\int\!w(t,n-1,x')\,\nu\,\pi(t,x',x)\,B(t,x,x')\dd{x'}-\int\!w(t,n,x)\,\nu\,\pi(t,x,x')\,\dd{x'}\\
    &\qquad-\lambda(t,x)\,w(t,n,x)
  \end{aligned}
\end{equation}
where $\nu>0$ is the intensity of homogenous Poisson process and
\begin{equation*}
  \begin{gathered}
    A(t,x)=\int{\alpha(t,x,x')\dd{x'}}, \qquad
    \pi(t,x,x')=\frac{\alpha(t,x,x')}{A(t,x)},\\
    B(t,x,x')=\frac{A(t,x)}{\nu}, \qquad
    \lambda(t,x)=A(t,x)-\nu.
  \end{gathered}
\end{equation*}
Note that $\pi$ is the probability kernel of the embedded chain $\rdm{\hat{X}}$ and $A(t,x)$ is the intensity of the $\rdm{T}^X_k$ process.
We recognize this equation as the filter equation with the rate-$\nu$ Poisson-process driver, boost $B$, and decay $\lambda$.

>>>>

Denote the underlying probability space by $(\Omega,\Borel,\mathrm{Prob})$.
We will assume that our population process is a time-inhomogeneous Markov jump process, $\X_t$, parameterized by time $t\in\Rp\colonequals\{t\in\mathbb{R}\;\vert\;t\ge{0}\}$ and taking values in some space $\Xspace$.
In earlier work \citep{King2022}, we limited ourselves to the case $\Xspace=\Z^d$, but here we assume only that $\Xspace$ is a complete metric space with a countable dense subset, \ie a Polish space.
The population process is completely specified by its initial-state distribution, $p_0$, and its transition rates $\alpha$.
In particular, we suppose that
\begin{equation}\label{eq:ic}
  \Prob{\X_0\in\mathcal{E}}=\int_{\mathcal{E}}{p_0(x)\,\dd{x}}
\end{equation}
for all measurable sets $\mathcal{E}\subseteq\Xspace$.
For any $t\in\Rp$, $x,x'\in\Xspace$, we think of the quantity $\alpha(t,x,x')$ as the instantaneous hazard of a jump from $x$ to $x'$.
More precisely, the transition rates have the following properties:
\begin{equation*}
  \begin{gathered}
    \alpha(t,x,x')\ge{0}, \qquad \int_{\Xspace}{\alpha(t,x,x')\,\dd{x'}}<\infty,\\
  \end{gathered}
\end{equation*}
for all $t\in\Rp$ and $x,x'\in\Xspace$.
Henceforth, we understand that integrals are taken over all of $\Xspace$ unless otherwise specified.
Let $\N_t$ be the number of jumps that $\X$ has taken by time $t$.
We assume that $\N_t$ is a simple counting process so that
\begin{equation*}
  \begin{gathered}
    \Prob{\N_{t+\Delta}=n+1\;\vert\;\N_{t}=n}=\Delta\,\int{\alpha(t,x,x')\,\dd{x'}}+o(\Delta),
    \qquad
    \Prob{\N_{t+\Delta}>n+1\;\vert\;\N_{t}=n}=o(\Delta),\\
    \Prob{\X_{t+\Delta}\in\mathcal{E}\;\vert\;\X_{t}=x, \N_{t+\Delta}-\N_{t}=1}=\frac{\int_{\mathcal{E}}{\alpha(t,x,x')\,\dd{x'}}}{\int{\alpha(t,x,x')\,\dd{x'}}}.
  \end{gathered}
\end{equation*}
We will further assume that $\X_t$ is non-explosive, \ie, that $\Prob{\N_t<\infty}=1$ for all $t$.
\AAK{[Is this equivalent to non-explosivity? Or merely an implication?]}

The above may be compactly summarized by stating that if $w(t,x)$ satisfies the Kolmogorov forward equation (KFE),
\begin{equation}\label{eq:kfe}
  \frac{\partial{w}}{\partial{t}}(t,x)=\int\!w(t,x')\,\alpha(t,x',x)\,\dd{x'}-\int\!w(t,x)\,\alpha(t,x,x')\,\dd{x'},
\end{equation}
and if, moreover, $w(0,x) = p_0(x)$,
then $\int_{\mathcal{E}}{w(t,x)\,\dd{x}}=\Prob{\X_t\in\mathcal{E}}$ for every measurable $\mathcal{E}\subseteq{\Xspace}$.
\Cref{eq:kfe} is sometimes called the \emph{master equation} for $\X_t$.

@@

Next, we derive expressions for the likelihood of pruned and obscured genealogies, unconditional on the history.
For this, we use the filter equation technology developed in the Appendix.
In particular, the following theorem follows immediately from \cref{prop:fullfilt}.

\begin{thm}
  Suppose we have a pruned genealogy $P^*_T$, $T>0$, with event times $\event(P^*_T)$.
  Let $y^*=Y^{\Prune^*_T}$.
  For $x,x'\in\Xspace$, $y,y'\in\Yspace$, $u\in\Jumps$, define
  \begin{equation*}
    \begin{gathered}
      \phi_u\big(x,y,y')=\BinRatio{n(x)}{\ell(y')}{r^u}{s(y,y')}\qquad
      \text{and} \qquad
      \beta_u(t,x,x')=\alpha_u(t,x,x')\,Q_u(y,y'),
    \end{gathered}
  \end{equation*}
  where $Q_u(y,y')$ is as in \cref{thm:pruned_lik}.
  Fix $\nu>0$ and suppose $w=w(t,x)$ satisfies the filter equation
  \begin{equation}
    \begin{aligned}
      \frac{\partial{w}}{\partial{t}}=
      &\sum_u{\int{w(t,x')\,\beta_u(t,x',x)\,\phi_u\big(x,\yt^*_t,y^*_t\big)\,\dd{x'}}}
      -\sum_u{\int{w(t,x)\,\beta_u(t,x,x')\,\dd{x'}}}\\
      &\quad+\sum_{e\in\event(P^*_T)}{\delta(t,e)\,\left\{\sum_u{\int{w(t,x')\,\frac{\beta_u(t,x',x)}{\nu}\,\phi_u\big(x,\yt^*_t,y^*_t\big)\,\dd{x'}}}-w(t,x)\right\}},
    \end{aligned}
  \end{equation}
  with the initial condition $w(0,x)=p_0(x)$.
  Then the likelihood of $P^*_T$ is
  \begin{equation*}
    \lik(P^*_T)=\int{w(T,x)\,\dd{x}}.
  \end{equation*}
\end{thm}

@@

\appendix
\setcounter{thm}{0}
\setcounter{equation}{0}
\renewcommand{\theequation}{\thesection\arabic{equation}}

@@

Explicit expressions for the quantities that arise in this paper are not always readily available.
Here, we develop tools for manipulating complex expressions that are otherwise cumbersome.

\begin{defn}
  Suppose $\X_t$ is a continuous-time Markov process with Kolmogorov forward equation (KFE)
  \begin{equation}\label{eq:kfe2}
    \frac{\partial{w}}{\partial{t}}
    =\int{w(t,x')\,\beta(t,x',x)\,\dd{x'}}
    -\int{w(t,x)\,\beta(t,x,x')\,\dd{x'}}.
  \end{equation}
  Suppose that $B(t,x,x')>0$ and $\lambda(t,x)\in\R$ are given measurable functions.
  Let $\mathcal{M}\subseteq\Rp$ be countable such that $\mathcal{M}\cap[0,t]$ is finite for all $t$.
  Let $\nu>0$.
  We say that the equation
  \begin{equation}\label{eq:filtereq}
    \begin{aligned}
      \frac{\partial{w}}{\partial{t}}
      =&\int{w(t,x')\,\beta(t,x',x)\,B(t,x',x)\,\dd{x'}}
      -\int{w(t,x)\,\beta(t,x,x')\,\dd{x'}}
      -\lambda(t,x)\,w(t,x)\\
      &\quad+\sum_{e\in\mathcal{M}}\left\{\int{w(t,x')\,\frac{\beta(t,x',x)}{\nu}\,B(t,x',x)\,\dd{x'}}-w(t,x)\right\}.
    \end{aligned}
  \end{equation}
  is the \emph{filter equation} with \emph{driver} $\X_t$, \emph{boost} $B$, \emph{decay} $\lambda$, and \emph{observed events} $\mathcal{M}$, \emph{relative to} $\nu$.
\end{defn}

\begin{remark}
  \Cref{eq:filtereq} is equivalent to
  \begin{equation*}
    \begin{gathered}
      \frac{\partial{w}}{\partial{t}}
      =\int{w(t,x')\,\beta(t,x',x)\,B(t,x',x)\,\dd{x'}}
      -\int{w(t,x)\,\beta(t,x,x')\,\dd{x'}}
      -\lambda(t,x)\,w(t,x),\quad t\notin{\mathcal{M}},\\
      w(t,x)=\int{\leftlim{w}(t,x')\,\frac{\beta(t,x',x)}{\nu}\,B(t,x',x)\,\dd{x'}},\quad t\in{\mathcal{M}}.
    \end{gathered}
  \end{equation*}
\end{remark}

Filter equations afford a convenient means of computing expectations and likelihoods for pure jump processes.
This is facilitated by the following fact.

\begin{lemma}\label{lemma:filtereq}
  The filter equation \eqref{eq:filtereq} with $\mathcal{M}=\emptyset$ is satisfied by $w(t,x)=\int_0^{\infty}{v\,u(t,x,v)\,\dd{v}}$, where $u$ satisfies the KFE
  \begin{equation}\label{eq:filterlemma}
    \begin{aligned}
      \frac{\partial{u}}{\partial{t}}=&\int{u(t,x',v')\,\beta(t,x',x)\,\delta(v,B(t,x',x)\,v')\,\dd{x'}\,\dd{v'}}\\
      &\qquad-\int{u(t,x,v)\,\beta(t,x,x')\,\delta(v',B(t,x,x')\,v)\,\dd{x'}\,\dd{v'}}
      +\tfrac{\partial}{\partial{v}}\left[\lambda(t,x)\,v\,u(t,x,v)\right],
    \end{aligned}
  \end{equation}
  on the space $\Xspace\times(0,\infty)$.
  Here, $\delta(v,v')$ is the familiar Dirac $\delta$.
\end{lemma}
\begin{proof}
  \begin{equation*}
    \begin{aligned}
      \frac{\partial{w}}{\partial{t}}
      =&\int{v\,\frac{\partial{u}}{\partial{t}}(t,x,v)\,\dd{v}}\\
      =&\int{v\,u(t,x',v')\,\beta(t,x',x)\,\delta(v,B(t,x',x)v')\,\dd{v}\,\dd{x'}\,\dd{v'}}\\
      &\qquad-\int{v\,u(t,x,v)\,\beta(t,x,x')\,\delta(v',B(t,x,x')v)\,\dd{v}\,\dd{x'}\,\dd{v'}}\\
      &\qquad+\int{v\,\tfrac{\partial}{\partial{v}}\left[\lambda(t,x)\,v\,u(t,x,v)\right]\,\dd{v}}.\\
    \end{aligned}
  \end{equation*}
  Evaluating the first integral with respect to $v$, the second with respect to $v'$, and the third by parts, we obtain
  \begin{equation*}
    \begin{aligned}
      \frac{\partial{w}}{\partial{t}}
      =&\int{v'\,u(t,x',v')\,\beta(t,x',x)\,B(t,x',x)\,\dd{v'}\,\dd{x'}}
      -\int{v\,u(t,x,v)\,\beta(t,x,x')\,\dd{v}\,\dd{x'}}\\
      &\qquad-\lambda(t,x)\,\int{v\,u(t,x,v)\,\dd{v}},
    \end{aligned}
  \end{equation*}
  which is simplified to obtain \cref{eq:filtereq}.
\end{proof}

\Cref{eq:filterlemma} is recognizable as the KFE of a certain process $(\X_t,\rdm{V}_t)$.
In particular, $\X_t$ is the driver with KFE \eqref{eq:kfe2}.
The $\rdm{V}_t$ is \emph{directed} by $\X_t$ in the sense that $\rdm{V}$ has jumps wherever $\X$ does:
when $\X$ jumps at time $t$ from $x$ to $x'$, $\rdm{V}$ jumps by the multiplicative factor $B(t,x,x')>0$.
Between jumps, $\rdm{V}_t$ decays deterministically and exponentially at rate $\lambda(t,x)$.
If we view $V_t$ as a weight, then \cref{lemma:filtereq} tells us how the $\rdm{V}_t$-weighted average of $\rdm{X}_t$ evolves in time:
this average is simply $\int{w(t,x)\,\dd{x}}$.
This motivates the following result, which shows how filter equations allow one to integrate over random histories.

\begin{prop}\label{prop:zeroboost}
  Suppose $\X_t$ is a non-explosive pure jump process with KFE
  \begin{equation*}
    \begin{gathered}
      \frac{\partial{w}}{\partial{t}}(t,x)=\int{w(t,x')\,\alpha(t,x',x)\,\dd{x'}}
      -\int{w(t,x)\,\alpha(t,x,x')\,\dd{x'}},\qquad
      w(0,x)=p_0(x).
    \end{gathered}
  \end{equation*}
  Let $\Hist_t$ be its history process.
  Suppose $\rdm{V}_t$ is a real-valued random process such that
  \begin{equation*}
    \Expect{\rdm{V}_t\;\vert\;\Hist_t=H_t}=\prod_{e\in\event(H_t)}{Q(e,\leftlim{X}_e,X_e)\,B(e,\leftlim{X}_e,X_e)},
  \end{equation*}
  for some given measurable functions $B>{0}$, $Q\in\{0,1\}$.
  If $w$ satisfies the filter equation,
  \begin{equation*}
    \begin{aligned}
      \frac{\partial{w}}{\partial{t}}(t,x)
      =&\int{w(t,x')\,\alpha(t,x',x)\,Q(t,x',x)\,B(t,x',x)\,\dd{x'}}
      -\int{w(t,x)\,\alpha(t,x,x')\,Q(t,x,x')\,\dd{x'}}\\
      &\qquad-\int{w(t,x)\,\alpha(t,x,x')\,\Big(1-Q(t,x,x')\Big)\,\dd{x'}},
    \end{aligned}
  \end{equation*}
  then $\Expect{\rdm{V}_t}=\int{w(t,x)\,\dd{x}}$.
\end{prop}
\begin{proof}
  Apply \cref{lemma:filtereq} with $\beta(t,x,x')=\alpha(t,x,x')\,Q(t,x,x')$.
\end{proof}

%% An important special case is that of a deterministic driving process.
%% The following result is established by routine calculation.

%% \begin{prop}
%%   Suppose $X:[0,T]\to\Xspace$ is a deterministic, piecewise constant, \cadlag\ function and let $M=\event(X)$.
%%   Then the KFE for $X_t$ is \cref{eq:kfe2} with $\beta(t,x,x')=\sum_{e\in{M}}{\delta(t,e)\delta(x',X_e)}$.
%%   With this driver, the filter equation (\cref{eq:filtereq}) becomes
%%   \begin{equation*}
%%     \begin{gathered}
%%       \frac{\partial{w}}{\partial{t}}=
%%       -\lambda(t,x)\,w(t,x)\quad\text{for}\quad{t\in[0,T]\setminus{M}},\\
%%       w(t,x)=\delta(x,X_t)\,\int{\wt(t,x')\,B(t,x',X_t)\,\dd{x'}}\quad\text{for}\quad{t\in{M}}.
%%     \end{gathered}
%%   \end{equation*}
%% \end{prop}

Filter equations allow us to pass easily between equivalent representations of a process.
For example, an equivalent way of representing $\X_t$ is in terms of its embedded chain and event times.
Let $\rdm{\hat{X}}_k$ be the embedded chain of $\X_t$ and let $\rdm{\hat{T}}_k$ be the point process of its event times.
It is elementary that
\fontsize{10pt}{12pt}\selectfont
\begin{equation*}
  \begin{gathered}
    \Prob{\rdm{\hat{X}}_k=\hat{X}_k\;\Big\vert\;\rdm{\hat{X}}_{k-1}=\hat{X}_{k-1},\rdm{\hat{T}}_k=\hat{T}_k}=\frac{\alpha(\hat{T}_k,\hat{X}_{k-1},\hat{X}_k)}{\int{\alpha(\hat{T}_k,\hat{X}_{k-1},x')\dd{x'}}},\\
    \Prob{\rdm{\hat{T}}_k>\rdm{\hat{T}}_{k-1}+t\;\Big\vert\;\rdm{\hat{X}}_{k-1}=\hat{X}_{k-1},\rdm{\hat{T}}_{k-1}=\hat{T}_{k-1}}=\exp{\left(-\int_{0}^{t}{\int{\alpha(\hat{T}_{k-1}+s,\hat{X}_{k-1},x')\,\dd{x'}}\,\dd{s}}\right)}.
  \end{gathered}
\end{equation*}
\normalfont
Making the definitions,
\begin{equation}\label[pluralequation]{eq:nudefs}
  \begin{gathered}
    A(t,x)=\int{\alpha(t,x,x')\,\dd{x'}}, \qquad
    \pi(t,x,x')=\frac{\alpha(t,x,x')}{A(t,x)},\\
    B(t,x,x')=\frac{A(t,x)}{\nu}, \qquad
    \lambda(t,x)=A(t,x)-\nu,
  \end{gathered}
\end{equation}
we can rewrite the KFE as
\begin{equation}\label{eq:poissdriver}
  \frac{\partial{w}}{\partial{t}}(t,x)=
  \int{w(t,x')\,\nu\,\pi(t,x',x)\,B(t,x,x')\,\dd{x'}}
  -\int{w(t,x)\,\nu\,\pi(t,x,x')\,\dd{x'}}
  -\lambda(t,x)\,w(t,x).
\end{equation}
Here, $\nu>0$ is the intensity of a time-homogenous Poisson process.
Note that $\pi$ is the probability kernel of the embedded chain $\rdm{\hat{X}}$ and $A(t,x)$ is the intensity of the $\rdm{\hat{T}}_k$ process.
We recognize this equation as the filter equation with boost $B$, decay $\lambda$, and driver generated by $\nu\,\pi(t,x,x')$.
It corresponds to the following procedure for simulating $\X_t$:
\begin{compactenum}[(a)]
\item Simulate jump times according to the rate-$\nu$ Poisson process.
\item Simulate the embedded chain $\rdm{\hat{X}}_k$ using the kernel $\pi$.
\item Weight the realization by the product of the $B$ factors.
  Note that this makes the appropriate importance-sampling correction.
\end{compactenum}

By integrating equations of this kind, we effectively compute expectations with respect to the $\X_t$ process.
In particular, we can use \cref{prop:zeroboost} or \cref{eq:poissdriver} to integrate out the unobserved event times $\rdm{\hat{T}}_k$, as well as the unobserved states $\rdm{\hat{X}}_k$.
However, we also need to compute likelihoods for data that include observed event times.
Technically, in such a case, the likelihood is a density on the cone $\mathbb{C}_t=\coprod_{m=0}^{\infty}{\mathbb{C}^m_t}$, where $\mathbb{C}^m_t=\{t\in\halfclosed{0,t}\;\vert\;t_1<t_2<\dots<t_m\}$.
To avoid having to work with such complex objects, it is convenient to represent the likelihood as a density with respect to a constant-rate Poisson point process.
This is accomplished as follows.
